Redirecting Resolution Proofs

### 2011-05-10 bib/prooftransform/redirecting-proofs-2/paper.tex
wrote "The Algorithm" section and sketched a conclusion

Direct proofs can be done the same way, but they have the constraints that we
may not assume the negated conjecture $\B{0}$ in any of the sequent and that the
last sequent has the conjecture $0$ as succedent. However, in some of the direct
proofs, we will find it useful to introduce case splits, as we had in the Isar
proof above. For example:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\frm 1, 2 \hav 4 \end
\frx 3 \hen 5 \end
\frm \hav 6, 7 \end
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&6] \\
      4 \hence {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&7] \\
      5 \hence {} & 0
    \end{aligned}$
  \end{tabular}}
\end{tabular}}\]
The notation in brace is a case split. In general, a case split has the form
\[\left[\begin{tabular}{l|c|l}
  $\begin{aligned}[t]
    [&\Gamma_1] \\
     \Delta_{11} \have {} & \mathrm{E}_{11} \\
     & ~\vdots \\
     \Delta_{1k_1} \have {} & \mathrm{E}_{1k_1}
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    ~\ldots~ \\
    ~\ldots~ \\
	\\
    ~\ldots~
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    [&\Gamma_n] \\
     \Delta_{n1} \have {} & \mathrm{E}_{n1} \\
     & ~\vdots \\
     \Delta_{nk_n} \have {} & \mathrm{E}_{nk_n}
    \end{aligned}$
\end{tabular}
\right]\]
with the requirement that a sequent with the succedent $\Gamma_1,
\ldots, \Gamma_n$ has been proved already. Also, each of the branches
must be a valid proof. The assumptions $[\Gamma_{\!j}]$ may be used to
discharge assumptions in the same branch, just as if they had been
sequents ${} \have \Gamma_{\!j}$. Seen from outside, the case split
expression stands for a sequent with the succedent
$\mathrm{E}_{1k_1}, \ldots, \mathrm{E}_{nk_n}$.

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
[2] & [&5] & [6] \\
    & $5 \have {}$ & $2 \lor 4$
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more aesthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.

This gives us a new case split, but notice that 6 is dominated by 2, 3, and 4.
We start with it:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|l|l|r@{}l}
[2] & [3] & [4] & [&6] \\
& & & $6 \have {}$ & $3 \lor 4$
\end{tabular}}\end{tabular}}\]
This proves $2 \lor 3 \lor 4$. Since all but one branches are trivial, we
abbreviate it:
\[2 \lor 3 \lor 4 \lor 6 \have 2 \lor 3 \lor 4\]
It might help to think of such steps as ``rewriting'' steps, where
6 is rewritten into $3 \lor 4$. Now, 4 is dominated by 2 and 3, so we perform
another case split:
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|l|r@{}l}
[2] & [3] & [&4] \\
& & $4 \have {}$ & $2 \lor 3$
\end{tabular}}\end{tabular}}\]
And we abbreviate it:
\[2 \lor 3 \lor 4 \have 2 \lor 3\]
We are left with the conclusion $2 \lor 3$. The rest is reminiscent of our
second lasso-shaped proof:
\[\hbox{\begin{tabular}{@{}p{.25\textwidth}@{}p{.25\textwidth}@{}}
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&2] \\
      2 \have {} & 1
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&3] \\
      3 \have {} & 1
    \end{aligned}$
  \end{tabular}}
\frm 1 \hav 0 \end
\end{tabular}}\]
%
Gluing all of this together (and using $\hencesym$), we get

@@

There is one more concept that needs to be explained: that of a
``target''. The target set is a set of formulas whose disjunction is
what we want to prove. For the top-level proof, the target is the
singleton consisting only of the conjecture. For subproofs, the target

Our approach is based on four basic tools: contraposition (on
sequents), cut, case split, and ``rewriting.'' Before we look at the
actual algorithm, it will help to see these tools in action.

Cut (our second tool) is implicit in this notation. The proof
tree makes it explicit:
%
\begin{prooftree}
\AXC{\phantom{.}}
\UIC{${} \have 2$\strut}
  \AXC{\phantom{.}}
  \UIC{$2 \have 1$\strut}
\RL{\textsc{Cut}}
\BIC{${} \have 1$\strut}
  \AXC{\phantom{.}}
  \UIC{$1 \have 0$\strut}
\RL{\textsc{Cut}}
\BIC{${} \have 0$\strut}
\end{prooftree}

In general, reordering might not suffice. For example, for the proof
by contradiction used as the running example in
Section~\ref{sec:proof-notations}, the contrapositive sequents are
%
\begin{align*}
1, 2 & \have 4 \\
3, 4 & \have 5 \\
4, 6 & \have 0 \\
5, 7 & \have 0 \\
{} & \have 6, 7
\end{align*}

We can start outputing the proof
%
\begin{align*}
1, 2 & \have 4 \\
3, 4 & \have 5 \\
{} & \have 6, 7
\end{align*}
%
but then we get stuck: There is no obvious way to use the remaining
two sequents, $4, 6 \have 0$ and $5, 7 \have 0$, because we only know
``$6$ or $7$.'' With a case split (our third tool), we can finish the
proof. In the left branch, we assume $6$, which allows us to use the
sequent $4, 6 \have 0$ to deduce $0$, and similarly in the right
branch. We saw the full proof in Section~\ref{sec:proof-notations}.
Case split can be seen as a special case of a multi-cut rule.

Alternative: Using ``rewriting'' (our fourth tool), the sequent ${}
\have 6, 7$ can be transformed until we reach the desired conclusion.
E.g.
%
\begin{align*}
4 & \have 0, 7 \\
4, 5 & \have 0
\end{align*}

This is justified by the following proof tree:
%
\begin{prooftree}
\AXC{${} \have 6, 7$\strut}
  \AXC{$4, 6 \have 0$\strut}
\RL{\textsc{Cut}}
\BIC{$4 \have 0, 7$\strut}
  \AXC{$5, 7 \have 0$\strut}
\RL{\textsc{Cut}}
\BIC{$4, 5 \have 0$\strut}
\end{prooftree}
%
However, in this case, we prefer the case split, because it more
clearly reflects the structure of the underlying proof. As the proof
tree indicates, rewriting is again an instance of cut.

@@

  * method is not terribly difficult
  * and if would not surprise us if it was part of folklore or
    some special case of some existing work

In a second step, we are interested in transformations that increase
proof readability, such as those available for Mizar proofs
\cite{pak-2010}.

\paragraph{Acknowledgment.}
Tobias Nipkow.
Geoff Sutcliffe and Joseph Urban motivated me to look into this.
Andrei Popescu, with his deep theoretical insights, helped simplify my
previous framework.

>>>>

Direct proofs can be done the same way, but they have the constraints that we
may not assume the negated conjecture $\B{0}$ in any of the sequent and that the
last sequent has the conjecture $0$ as succedent. However, in some of the direct
proofs, we will find it useful to introduce case splits, as we had in the Isar
proof above. For example:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\frm 1, 2 \hav 4 \end
\frx 3 \hen 5 \end
\frm \hav 6, 7 \end
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&6] \\
      4 \hence {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&7] \\
      5 \hence {} & 0
    \end{aligned}$
  \end{tabular}}
\end{tabular}}\]
The notation in brace is a case split. In general, a case split has the form
\[\left[\begin{tabular}{l|c|l}
  $\begin{aligned}[t]
    [&\Gamma_1] \\
     \Delta_{11} \have {} & \varphi_{11} \\
     & ~\vdots \\
     \Delta_{1k_1} \have {} & \varphi_{1k_1}
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    ~\ldots~ \\
    ~\ldots~ \\
	\\
    ~\ldots~
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    [&\Gamma_n] \\
     \Delta_{n1} \have {} & \varphi_{n1} \\
     & ~\vdots \\
     \Delta_{nk_n} \have {} & \varphi_{nk_n}
    \end{aligned}$
\end{tabular}
\right]\]
with the requirement that a sequent with the succedent $\Gamma_1,
\ldots, \Gamma_n$ has been proved already. Also, each of the branches
must be a valid proof. The assumptions $[\Gamma_{\!j}]$ may be used to
discharge assumptions in the same branch, just as if they had been
sequents ${} \have \Gamma_{\!j}$. Seen from outside, the case split
expression stands for a sequent with the succedent
$\varphi_{1k_1}, \ldots, \varphi_{nk_n}$.

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
 & [&5] & \\
{[2]} & $5 \have {}$ & $2 \lor 4$ & [6]
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more aesthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.

This gives us a new case split, but notice that 6 is dominated by 2, 3, and 4.
We start with it:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|l|l|r@{}l}
 &  & & [&6] \\
{[2]} & [3] & [4] & $6 \have {}$ & $3 \lor 4$
\end{tabular}}\end{tabular}}\]
This proves $2 \lor 3 \lor 4$. Since all but one branches are trivial, we
abbreviate it:
\[2 \lor 3 \lor 4 \lor 6 \have 2 \lor 3 \lor 4\]
It might help to think of such steps as ``rewriting'' steps, where
6 is rewritten into $3 \lor 4$. Now, 4 is dominated by 2 and 3, so we perform
another case split:
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|l|r@{}l}
 & & [&4] \\
{[2]} & [3] & $4 \have {}$ & $2 \lor 3$
\end{tabular}}\end{tabular}}\]
And we abbreviate it:
\[2 \lor 3 \lor 4 \have 2 \lor 3\]
We are left with the conclusion $2 \lor 3$. The rest is reminiscent of our
second lasso-shaped proof:
\[\hbox{\begin{tabular}{@{}p{.25\textwidth}@{}p{.25\textwidth}@{}}
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&2] \\
      2 \have {} & 1
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&3] \\
      3 \have {} & 1
    \end{aligned}$
  \end{tabular}}
\frm 1 \hav 0 \end
\end{tabular}}\]
%
Putting all of this together (and using $\hencesym$), we get

@@

The process we applied to the examples in the previous section can be
generalized into an algorithm. The algorithm relies on the proof by
contradiction expressed as a set of sequents, which are turned around using the
contrapositive so that the formulas are unnegated (as we did in all the
examples). It maintains a set of \emph{proved nodes} (initially the empty set)
and a set of \emph{target nodes} that it may proceed to prove (initially the set
of all unnegated nodes). The constructed proof is expressed using the shorthand
notation.

\begin{enumerate}
\item Derive as many sequents as possible with their conclusion in the target set
based on the proved nodes. Each time a sequent is appended to the proof, its
conclusion is added to the set of proved nodes.

\item If all the nodes in the target set are proved, we are done. Otherwise, the
last sequent must be of the form $\Gamma \have c_1 \lor \cdots \lor c_n$ for $n
\ge 2$. Perform an $n$-way case split:%
\footnote{A generalization would be to perform an $m$-way case split, with $m <
n$, on disjuncts. For example, we could do a 3-way case split with $c_1 \lor
c_2$, $c_3$, and $c_4$ as the assumptions instead of breaking all the
disjunctions and doing a 4-way split. This could potentially lead to cleaner
proofs, if the groupings are chosen carefully. We will not explore this avenue
further in this paper.}
\[
\left[\begin{tabular}{c|c|c}
$[c_1]$ & $\cdots$ & $[c_n]$ \\
 & & 
\end{tabular}
\right]
\]

\item For each of the branches, invoke the procedure recursively, with the
assumption added to the proved set and with all the nodes that dominate the
branch and no other branch as the target set.

\item After the case split, go back to step 1. The case split itself is seen as
a sequent of the form $\Gamma \have c_1 \lor \cdots \lor c_n$ as far as step 2
is concerned.
\end{enumerate}

Once this procedure has terminated, we abbreviate case splits in which all
but one branch are trivial using ``rewriting,'' transforming
\[
\left[\begin{tabular}{c|c|c|r@{}l|c|c|c}
&&& $[$ & $c_i]$ &&& \\
&&& $\Delta_1 \have {}$ & $d_1$ &&& \\
&&&& $\vdots$ &&& \\
$[c_1]$ & $\cdots$ & $[c_{i-1}]$ & $\Delta_n \have {}$ & $d_n$ & $[c_{i+1}]$ & $\cdots$ & $[c_m]$
\end{tabular}
\right]
\]
into
\begin{align*}
\Delta_1 \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_1 \lor c_{i+1} \lor \cdots c_m \\
& \vdots \\
\Delta_n \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_n \lor c_{i+1} \lor \cdots c_m \\
\end{align*}
This works even if the $c_i$ branch has case splits. What we are doing
effectively is taking the nontrivial branch and adding $c_1, \ldots, c_{i-1},
c_{i+1}, \ldots, c_m$ as disjuncts in the conclusions.

Finally, we clean up the proof by introducing $\hencesym$ and translate it to
Isar.

%There is one more concept that needs to be explained: that of a
%``target''. The target set is a set of formulas whose disjunction is
%what we want to prove. For the top-level proof, the target is the
%singleton consisting only of the conjecture. For subproofs, the target
%
%Our approach is based on four basic tools: contraposition (on
%sequents), cut, case split, and ``rewriting.'' Before we look at the
%actual algorithm, it will help to see these tools in action.
%
%Cut (our second tool) is implicit in this notation. The proof
%tree makes it explicit:
%%
%\begin{prooftree}
%\AXC{\phantom{.}}
%\UIC{${} \have 2$\strut}
%  \AXC{\phantom{.}}
%  \UIC{$2 \have 1$\strut}
%\RL{\textsc{Cut}}
%\BIC{${} \have 1$\strut}
%  \AXC{\phantom{.}}
%  \UIC{$1 \have 0$\strut}
%\RL{\textsc{Cut}}
%\BIC{${} \have 0$\strut}
%\end{prooftree}
%
%In general, reordering might not suffice. For example, for the proof
%by contradiction used as the running example in
%Section~\ref{sec:proof-notations}, the contrapositive sequents are
%%
%\begin{align*}
%1, 2 & \have 4 \\
%3, 4 & \have 5 \\
%4, 6 & \have 0 \\
%5, 7 & \have 0 \\
%{} & \have 6, 7
%\end{align*}
%
%We can start outputing the proof
%%
%\begin{align*}
%1, 2 & \have 4 \\
%3, 4 & \have 5 \\
%{} & \have 6, 7
%\end{align*}
%%
%but then we get stuck: There is no obvious way to use the remaining
%two sequents, $4, 6 \have 0$ and $5, 7 \have 0$, because we only know
%``$6$ or $7$.'' With a case split (our third tool), we can finish the
%proof. In the left branch, we assume $6$, which allows us to use the
%sequent $4, 6 \have 0$ to deduce $0$, and similarly in the right
%branch. We saw the full proof in Section~\ref{sec:proof-notations}.
%Case split can be seen as a special case of a multi-cut rule.
%
%Alternative: Using ``rewriting'' (our fourth tool), the sequent ${}
%\have 6, 7$ can be transformed until we reach the desired conclusion.
%E.g.
%%
%\begin{align*}
%4 & \have 0, 7 \\
%4, 5 & \have 0
%\end{align*}
%
%This is justified by the following proof tree:
%%
%\begin{prooftree}
%\AXC{${} \have 6, 7$\strut}
%  \AXC{$4, 6 \have 0$\strut}
%\RL{\textsc{Cut}}
%\BIC{$4 \have 0, 7$\strut}
%  \AXC{$5, 7 \have 0$\strut}
%\RL{\textsc{Cut}}
%\BIC{$4, 5 \have 0$\strut}
%\end{prooftree}
%%
%However, in this case, we prefer the case split, because it more
%clearly reflects the structure of the underlying proof. As the proof
%tree indicates, rewriting is again an instance of cut.

@@

We presented a method to transform proofs by contradiction as returned by
resolution provers into direct proofs. This sometimes introduces case splits,
but our procedure avoids duplicating proof steps in the different branches of
the split by joining again as early as possible. The result is direct Isar
proofs that have some of the structure Isabelle users have come to expect. Our
approach is fairly simple, it would not surprise us if it were part of folklore
or a special case of existing work.

In a second step, we are interested in transformations that increase proof
readability, such as those available for Mizar proofs \cite{pak-2010}. We also
need to address other issues not treated satisfactorily in Sledgehammer
\cite{paulson-susanto-2007}, notably Skolemization.

\paragraph{Acknowledgment.}
I want to thank Tobias Nipkow, Geoff Sutcliffe, and Joseph Urban for motivating
me to look into this. I am also thankful to Andrei Popescu, who with his deep
theoretical insights helped simplify the approach considerably.
